### 2025-02-23 paper/relatedwork.tex
Mention Naperian functors.

Automatic differentiation has been around for many decades~\cite{early-ad1, early-ad2},
so it is well-understood at a conceptual level.  However,
a number of questions related to bringing AD into the context of
programming languages remain open.  Recent successes in machine learning
have spurred further interest in AD which has led to several new developments.
For the context of this paper, we focus on recent work that contributes to 
the perspective of balancing correctness guarantees and performance.
Our selection here is by no means exhaustive, for
a broader scope we refer the reader to~\cite{autodiff-survey}.

@@

Verified high-performance computing is an area of active research, and
one line of research is based on the idea of separating the high level
\emph{specification} of a problem from the \emph{schedule} that
describe how it is to be executed. Verifying that such schedules are
semantics-preserving is the topic of several recent
publications~\cite{10.1145/3527328,10.1145/3498717}. Our approach
differs by not using a scheduling language, and instead focuses on
verifying the specification itself, and also by having a particularly
expressive specification language.

>>>>

\paragraph{Automatic Differentiation}
Automatic differentiation has been around for many decades~\cite{early-ad1, early-ad2},
so it is well-understood at a conceptual level.  However,
a number of questions related to bringing AD into the context of
programming languages remain open.  Recent successes in machine learning
have spurred further interest in AD which has led to several new developments.
For the context of this paper, we focus on recent work that contributes to 
the perspective of balancing correctness guarantees and performance.
Our selection here is by no means exhaustive, for
a broader scope we refer the reader to~\cite{autodiff-survey}.

@@

\paragraph{Verified high performance computing}
One popular line of research is based on the idea of separating the
high level \emph{specification} of a problem from the \emph{schedule}
that describe how it is to be executed. Verifying that such schedules
are semantics-preserving is the topic of several recent
publications~\cite{10.1145/3527328,10.1145/3498717}. Our approach
differs by not using a scheduling language, and instead focuses on
verifying the specification itself, and also by having a particularly
expressive specification language.

\paragraph{Typed rank polymorphism}
Gibbons showed how to express a form of rank polymorphism in Haskell,
through the use of \emph{Naperian
  functors}~\cite{10.1145/2976022.2976023}. This form of rank
polymorphism is somewhat more limited than ours; in particular, it
imposes static ranks. Further, Gibbons' work focused solely on the
embedding in Haskell, while our system also supports code extraction
and demonstrates good real-world performance.
